---
title: "covid_phylog"
author: "grp24"
date: "12/11/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---
#Set Up
libraries
```{r, echo=FALSE}
library(ape)
library(seqinr)
library(phangorn)
library(tidyverse)
library(ggtext)
library(RColorBrewer)
```

read in aligned data and grab some sequence info
```{r}
states<-read.aa("aligned_states_china_muscle.txt", format="fasta")
coding<-read.dna("aln-fasta.txt",format="fasta")
```

how to read in info from accession #s only
```{r, eval=F}
accession<-read.delim("accession_all.txt")
refseqR::save_AAfasta_from_xps(accession, "filepath/filename")
```


add in additional data
```{r}
new1<-read.delim("states_data.txt", header=FALSE)
new2<-read.delim("china_data.txt", header=FALSE)
new=rbind(new1,new2)
names(new)<-c("accession","state","date")
```

update labels to just be the accession number for easier graphing later on
```{r}
old=as.matrix(names(states))
states=updateLabel(states, old, as.character(new$accession))
```


#EDA
cleaning up data for use in plotting
```{r}
timeline<-new%>%
  na.omit()%>%
  filter(state!="GA"&state!="IN")%>% #take out incorrectly queried data
  mutate(dated= zoo::as.yearmon(date, "%Y-%m"))%>%
  mutate(dated=format(dated, "%m"))%>%
  mutate(period=ifelse(as.numeric(dated)<7,"pre July","post July"))%>%
  mutate(region=case_when(
    state %in% c("CA","NM","WA") ~ "West",
    state %in% c("FL","GA","TX","SC") ~ "South",
    state %in% c("NY","MD","VA","NC") ~ "East",
    state %in% c("MN","WI","IL","IN") ~ "Midwest",
    TRUE~ "China"
  ))%>%
  mutate(state=as.character(state))%>%
  mutate(state_date=paste(as.character(state),as.numeric(dated)))%>%
  na.omit()
```

distribution of sequence counts over time
```{r}
#change color pallete
palette(brewer.pal(n=8,name = "Dark2"))
timeline%>%ggplot(aes(dated,fill=region))+
  geom_bar()+
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=0.5),
        text=element_text(size=7))+
  labs(
    title = "**COVID spike protein**  
    <span style='font-size:11pt'> Sequence counts by US regions
    <span style='color:#A3A500;'>East</span>, 
    <span style='color:#00BF7D;'>Midwest</span>, 
    <span style='color:#00B0f6;'>South</span>, and
    <span style='color:#E76BF3;'>West</span>, including outgroup
    <span style='color:#F8766D;'>China</span>",
    subtitle="split into time intervals Pre July(n=79) and Post July(n=75)",
    x = "month (2020)", y = "sequence count"
  ) +
  geom_vline(xintercept = 6.5,color="darkgrey")+
  theme_minimal() +
  geom_text(x=6.3, y=34.7, label="pre July",col="darkgrey",size=3, angle=90)+
  geom_text(x=6.7, y=35, label="post July",col="darkgrey" ,size=3,angle=90)+
  theme(
    plot.title = element_markdown(lineheight = 1.1),
    legend.text = element_markdown(size = 11),
    legend.position = "none")
```


#Methods
change alignment to phyDat object for use in the phanghon package.
```{r}
states_phyDat <- phyDat(states, type = "AA", levels = NULL)
coding_phyDat<-phyDat(coding, type="DNA",levels=NULL)
```

run some model testing to see which distance matrix is best for our data.
We can use mt list to pick lowest AIC.
```{r}
#mt takes 4everrrr!! change cores to 2 for faster implem in parallel
#mt <- modelTest(states_phyDat, model="all") 
states_dist <- dist.ml(states, model="FLU")

coding_dist<-dist.dna(coding,model = "JC")
```

plotting AIC
```{r, eval=F}
mt%>%
  arrange(AIC)%>%
  ggplot(aes(reorder(Model,-AIC),AIC, col=AIC<8660))+
  geom_point()+
 scale_colour_manual(name = 'AIC > 8900', values = setNames(c('#1B9E77','black'),c(T, F)))+
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=0.5),
        text=element_text(size=7))+
  labs(
    title = "**AIC measures**  
    <span style='font-size:11pt'>for finding
    <span style='color:#1B9E77;'>best evolutionary models</span>
    to be used in distance matrix calculation",
    x = "Evolutionary Models", y = "AIC"
  )+
  theme_minimal() +
  theme(
    plot.title = element_markdown(lineheight = 1.1),
    legend.text = element_markdown(size = 11),
    axis.text.x=element_text(angle=60, hjust=1, vjust=1),
    legend.position = "none")

m2<-mt%>%
  ggplot(aes(reorder(Model,AIC),AIC))+
  geom_point(alpha=0.7)+
  scale_colour_manual(name = 'AIC > 8950', values = setNames(c('purple','black'),c(T, F)))+
  theme(axis.text.x=element_text(angle=60, hjust=1, vjust=1),
        text=element_text(size=7))+
  labs(
    title = "**AIC measures**  
    <span style='font-size:11pt'> for finding
    <span style='color:#F8766D;'>best</span> models</span>",
    x = "Model", y = "AIC"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(lineheight = 1.1))
  
```

tree construction for NJ, UPGMA, fastME
```{r}
states_UPGMA <- upgma(states_dist)
states_NJ  <- NJ(states_dist)
states_fastme<-fastme.bal(states_dist)

coding_UPGMA <- upgma(coding_dist)
coding_NJ  <- NJ(coding_dist)
coding_fastme<-fastme.bal(coding_dist)
```

plots coding region
```{r}
par(mfrow=c(1,3),oma = c(0, 0, 2, 0))
plot(coding_UPGMA,
     main = "UPGMA"
     ,type="cladogram",
     show.tip=TRUE)
tiplabels(pch=15,cex=.5, 
          col = as.factor(timeline$region[match(coding_UPGMA$tip.label, timeline$accession)])
          )
plot(coding_NJ,
     main = "NJ"
     ,type="cladogram",
     show.tip=FALSE)
tiplabels(pch=15,cex=.5, 
          col = as.factor(timeline$region[match(coding_UPGMA$tip.label, timeline$accession)])
          )
plot(coding_fastme,
     main = "fastME"
     ,type="cladogram",
     show.tip=FALSE)
tiplabels(pch=15,cex=.5, 
          col = as.factor(timeline$region[match(coding_fastme$tip.label, timeline$accession)])
          )
```


#Plots
region plots amino acids
```{r}
#for adding state labels
li<-timeline$state_date[match(states_UPGMA$tip.label, timeline$accession)]
#three plots
par(mfrow=c(1,3),oma = c(0, 0, 2, 0))

#UPGMA
plot(states_UPGMA,
     main = "UPGMA"
     ,type="cladogram",
     show.tip=FALSE)
tiplabels(pch=15,cex=.5, 
          col = as.factor(timeline$region[match(states_UPGMA$tip.label, timeline$accession)])
          )

#NJ
plot(states_NJ,
     main = "NJ",
     type="cladogram",
     show.tip=FALSE
     )
tiplabels(pch=15,cex=.5, 
          col = as.factor(timeline$region[match(states_NJ$tip.label, timeline$accession)])
          )

#fastME
plot(states_fastme,
     main = "FastME",
     type="cladogram",
     show.tip=FALSE
     )
tiplabels(pch=15, cex=.5, 
          col = as.factor(timeline$region[match(states_fastme$tip.label, timeline$accession)])
          )
```
region title with corresponding colors
```{r}
mtext(
  expression( 
    "Model selection based off of US region "*
    phantom("East ") * 
    phantom("Midwest ") * 
    phantom("South ") * 
    "and " *
    phantom( "West " )*
    "with outgroup "*
    phantom("China")
  ),
  outer = TRUE,cex=0.7, adj=0
  )
mtext(
  expression( 
    phantom("Model selection based off of US region ")*
    "East " *
    phantom("Midwest ") * 
    phantom("South ") * 
    phantom("and ") *
    phantom( "West " )*
    phantom("with outgroup ")*
    phantom("China")
  ),
  outer = TRUE,cex=0.7, adj=0, col=2
  )
mtext(
  expression( 
    phantom("Model selection based off of US region ")*
    phantom("East ") * 
    "Midwest " * 
    phantom("South ") * 
    phantom("and ") *
    phantom( "West " )*
    phantom("with outgroup ")*
    phantom("China")
  ),
  outer = TRUE,cex=0.7, adj=0, col=3
  )
mtext(
  expression( 
    phantom("Model selection based off of US region ")*
    phantom("East ") * 
    phantom("Midwest ") * 
    "South " * 
    phantom("and ") *
    phantom( "West " )*
    phantom("with outgroup ")*
    phantom("China")
  ),
  outer = TRUE,cex=0.7, adj=0, col=4
  )
mtext(
  expression( 
    phantom("Model selection based off of US region ")*
    phantom("East ") * 
    phantom("Midwest ") * 
    phantom("South ") * 
    phantom("and ") *
    "West "*
    phantom("with outgroup ")*
    phantom("China")
  ),
  outer = TRUE,cex=0.7, adj=0, col=5
  )
mtext(
  expression( 
    phantom("Model selection based off of US region ")*
    phantom("East ") * 
    phantom("Midwest ") * 
    phantom("South ") * 
    phantom("and ") *
    phantom("West ")*
    phantom("with outgroup ")*
    "China"
  ),
  outer = TRUE,cex=0.7, adj=0, col=1
  )
```

period plots
```{r}

#for adding state labels
li<-timeline$state_date[match(states_UPGMA$tip.label, timeline$accession)]

#three plots
par(mfrow=c(1,3),oma = c(0, 0, 2, 0))

#UPGMA
plot(states_UPGMA,
     main = "UPGMA"
     ,type="cladogram",
     show.tip=FALSE)
tiplabels(pch=15,cex=.5, 
          col = as.factor(timeline$period[match(states_UPGMA$tip.label, timeline$accession)])
          )

#NJ
plot(states_NJ,
     main = "NJ",
     type="cladogram",
     show.tip=FALSE
     )
tiplabels(pch=15,cex=.5, 
          col = as.factor(timeline$period[match(states_NJ$tip.label, timeline$accession)])
          )

#fastME
plot(states_fastme,
     main = "FastME",
     type="cladogram",
     show.tip=FALSE
     )
tiplabels(pch=15, cex=.5, 
          col = as.factor(timeline$period[match(states_fastme$tip.label, timeline$accession)])
          )
```
period title with corr colors
```{r}
#adding colored title
mtext(
  expression( 
    "Model selection based on time intervals "*
    phantom("Pre July ") * 
    " and " *
    phantom( "Post July" )
  ),
  outer = TRUE,cex=0.7, adj=0
  )
mtext(
  expression( 
    phantom("Model selection based on time intervals ")*
      "Pre July " * 
      phantom(" and " )*
      phantom( "Post July" )),
  outer = TRUE,cex=0.7, adj=0, col=2
  )
mtext(
  expression( 
    phantom("Model selection based on time intervals ")*
      phantom("Pre July ") * 
      phantom(" and " )*
      "Post July" ),
  outer = TRUE,cex=0.7, adj=0,col=1
  )
```

correlation plots
```{r}
x <- as.vector(states_dist)
y <- as.vector(as.dist(cophenetic(states_UPGMA)))
z <- as.vector(as.dist(cophenetic(states_NJ)))
v<-as.vector(as.dist(cophenetic(states_fastme)))

par(mfrow=c(3,1),mar=c(1,1,1,1))
plot(x, y, xlab="original pairwise distances", ylab="pairwise distances on the tree", main="UPGMA", pch=20, col="black", cex=2)
abline(lm(y~x), col=4)
cor(x,y)^2

plot(x, z, xlab="original pairwise distances", ylab="pairwise distances on the tree", main="NJ", pch=20, col="black", cex=2)
abline(lm(z~x), col=5)
cor(x,z)^2

plot(x, v, xlab="original pairwise distances", ylab="pairwise distances on the tree", main="fastME", pch=20, col="black", cex=2)
abline(lm(v~x), col=3)
cor(x,z)^2

```

#ML and Parsimony
```{r, eval=F}
fit = pml(states_NJ, states_phyDat)
fit
env <- attr(mt, "env")
fit <- eval(get("FLU", env), env)
fit
plotBS(midpoint(fit$tree), bs, p = 50, type="p")
```



#Junk Drawer
plotting states data
```{r}
library(usmap)
monthly<-timeline%>%group_by(state,date)%>%count()
periodly<-timeline%>%group_by(state,period)%>%count()
```
pick whether monthly or periodly
```{r}
p<-plot_usmap(data=periodly,values="n", 
           include = c("CA","NM","WA","FL","GA","TX",
                       "SC","NY","MD","VA","NC","MN",
                       "WI","IL","IN"),
           color="purple") +
  scale_fill_continuous(
    low = "white", high = "purple", name = "seq count", 
    label = scales::comma
  ) + theme(legend.position = "right")+
  labs(title="Sequence counts per month in 2020")+
  facet_wrap(~period)
```


  
